Score-based tests of measurement invariance: Use in practice

### 2013-10-02 frontiers/frontiers.Rnw
put spaces after commas.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@272 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor revisions to frontiers ms
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@273 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to Edgar Merkle. 
  Email: merklee@missouri.edu.}
\shorttitle{Measurement invariance test properties}
\rightheader{Measurement invariance test properties}

@@

Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in tests of measurement invariance.  BSEM
(Bayesian Structural Equation Modeling) generally handles these ideas by replacing across-group equality constraints with informative prior distributions.  In this paper, we describe an alternative approach: the development of test statistics that are sensitive to invariance violations of interest and insensitive to ``anomalous'' invariance violations.  The test statistics are specifically applicable to situations where one wishes to test for measurement invariance with respect to an ordinal variable, and they are special cases of a family of tests that may generally be used to study measurement invariance w.r.t.\ continuous, categorical, and ordinal variables.

The family of tests described in this paper have recently been studied in the context of measurement invariance \cite{MerZei13,MerFanZei}, though their practical application has been limited to a small set of simulations and datasets.  In this paper, we provide a detailed illustration of the tests' use in practice and a detailed examination 
of the tests' performance under scenarios likely to be encountered in practice. 
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios. 
In the following section, we first briefly review the 
theoretical framework of the proposed tests and
provide a short tutorial illustrating the use of the tests in R.
Next, we study the tests' performance in simulations that mimic practical research scenarios. 
Finally, we provide some further discussion on the tests' use in practice.

@@

\subsection{Continuous Treatment}
If we are interested in measurement invariance violations that are 
monotonic with the age groups, it is perhaps simplest to treat the 
age groups as continuous.  In doing so, we can use the statistics 
from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
That is, we can fit a model whose parameters 
are restricted to be equal across all individuals and then examine 
how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
age (where age ties are broken arbitrarily).
This is demonstrated below, with similar code being useful when one is 
testing for measurement invariance w.r.t.\ truly continuous variables.

@@

<<Get Estimate>>=
dm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "DM")
cvm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "CvM")
maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLM")
round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@
%% move this code description to the cat2BB part. 
%In the code above, the \verb+parm=1:4+ argument means that 
%we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
%which are supplied by lavaan.  The ordering of parameters can be obtained via the 
%command \verb+coef(reduced)+.  
We see that two of the three $p$-values output at 
the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

@@

Many of the applications in this volume, along with many applications in general, focus on testing for measurement invariance across unordered categories such as nations or gender.  As discussed earlier in this paper, the score-based tests for unordered categories are essentially equivalent to the usual likelihood ratio test.  Given a measurement invariance violation across these unordered categories, however, researchers typically wish to know why the violation occurred.  At this point, researchers may examine education level, socioeconomic status, income levels, and so on across the unordered categories.  These variables are often ordinal or continuous in nature, so that the family of tests described in this paper are applicable.  This is a first step towards describing why measurement invariance violations occur, as opposed to simply detecting measurement invariance violations. \readme{In the frontiers abstracts, one paper was proposed along these lines.  I will get a draft of that paper and then we can cite it here.}
The tests are convenient for this purpose, as they do not require a new model to be estimated for each ordinal variable.  Instead, each ordinal variable defines an ordering of observations, which in turn yields a test statistic that is specific to that ordinal variable.

>>>>

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to Edgar Merkle. 
  Email: merklee@missouri.edu.}
\shorttitle{Score-based tests in practice}
\rightheader{Score-based tests in practice}

@@

Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in corresponding tests of measurement invariance.  BSEM
(Bayesian Structural Equation Modeling) generally handles these ideas by replacing across-group equality constraints with informative prior distributions.  In this paper, we describe an alternative approach: the development of test statistics that are sensitive to invariance violations of interest and insensitive to ``anomalous'' invariance violations.  The test statistics are specifically applicable to situations where one wishes to test for measurement invariance with respect to an ordinal variable, and they are special cases of a family of tests that may generally be used to study measurement invariance w.r.t.\ continuous, categorical, and ordinal variables.

The family of tests described in this paper have recently been studied in the context of measurement invariance \cite{MerZei13,MerFanZei}, though their practical application has been limited to a small set of simulations and datasets.  In this paper, we provide a detailed illustration of the tests' use in practice and a detailed examination 
of the tests' performance under scenarios likely to be encountered in practice. 
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios. 
In the following sections, we first briefly review the 
theoretical framework of the proposed tests and
provide a short tutorial illustrating the use of the tests in R.
Next, we study the tests' performance in simulations that mimic practical research scenarios. 
Finally, we provide some further discussion on the tests' use in practice.

@@

\subsection{Continuous Treatment}
If we are interested in measurement invariance violations that are 
monotonic with the age groups, it is perhaps simplest to treat the 
age groups as continuous.  In doing so, we can use the statistics 
from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm} and~\eqref{eq:maxlm}.  
That is, we can fit a model whose parameters 
are restricted to be equal across all individuals and then examine 
how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
age (where age ties are broken arbitrarily).
This is demonstrated below, with similar code being useful when one is 
testing for measurement invariance w.r.t.\ truly continuous variables.

@@

<<getestimate>>=
dm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "DM")
cvm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "CvM")
maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLM")
round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@
%% move this code description to the cat2BB part. 
%In the code above, the \verb+parm=1:4+ argument means that 
%we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
%which are supplied by lavaan.  The ordering of parameters can be obtained via the 
%command \verb+coef(reduced)+.  
We see that two of the three $p$-values output at 
the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

@@

Many of the applications in this volume, along with many applications in general, focus on testing for measurement invariance across unordered categories such as nations or gender.  As discussed earlier in this paper, the score-based tests for unordered categories are essentially equivalent to the usual likelihood ratio test.  Given a measurement invariance violation across these unordered categories, however, researchers typically wish to know why the violation occurred.  At this point, researchers may examine education level, socioeconomic status, income levels, and so on across the unordered categories.  These variables are often ordinal or continuous in nature, so that the family of tests described in this paper are applicable.  This is a first step towards describing why measurement invariance violations occur, as opposed to simply detecting measurement invariance violations. \readme{In the abstracts for the special issue (which the special editors sent us), one paper was proposed along these lines.  I will get a draft of that paper and then we can cite it here.}
The tests are convenient for this purpose, as they do not require a new model to be estimated for each ordinal variable.  Instead, each ordinal variable defines an ordering of observations, which in turn yields a test statistic that is specific to that ordinal variable.
